---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

# Coursera Data Science Specialization: Reproducible Data - Activity Monitoring

[Introduction](#Intro)<br />
[Summary of Data](#Summary)<br />
[Repository Contents](#RepoStuff)<br/>
[Summary of Code and Analysis](#Code)<br />
[Conclusions](#Conclusions)<br/>
[Additional Information](#Other)<br />



<a name="Intro"/>
## Introduction

Technology has now made it possible to closely monitor an individual's health. Certain products such as [Fitbit](http://www.fitbit.com), [Nike
Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), and [Jawbone Up](https://jawbone.com/up) allow tracking of one's physical movement throughout the day. However, the raw data can be difficult to obtain and sometimes more difficult to understand, as there is a minimal amount of statistical analysis available as well as a lack of software for handling the data.

In this project, data from an activity monitoring device is used to provide insight into the typical movements of its user. Data is taken at 5 minute intervals through the day for the months of October 2012 and November 2012. In this analysis, we aim to determine what an average day looks like, and whether this varies during the week.



<a name="Summary"/>
## Summary of Data

The data for this project may be downloaded from [this page](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip). The dataset is a CSV-formatted file with a total of 17,568 observations in the dataset. The variables listed in this dataset are:

- **steps**: Number of steps taking in a 5-minute interval (missing values are `NA`)
- **date**: The date on which each measurement was taken (in YYYY-MM-DD format)
- **interval**: The identifier for each 5-minute interval during which the measurements were taken






<a name="RepoStuff"/>
## Repository Contents

- **doc** folder: Original .pdf of Coursera assignment
- **figure** folder: Figures produced by the analysis code, containing the following figures produced by the analysis described herein:
    - **Plot 1 - Histogram.png**
    - **Plot 2 - Mean Steps.png**
    - **Plot 3 - Histogram 2.png**
    - **Plot 4 - Mean Steps: Weekday vs. Weekend.png**
- **instructions_fig** folder: Figure produced by/for assignment instructions
- **Assignment_Description.md**: Markdown ReadMe version of Coursera assignment
- **PA1_template.md**: Markdown file template for analysis, including segments of code used for analysis
- **PA1_template.html**: HTML version of the markdown file
- **ReadMe.Rmd**: The ReadMe introduction and explanation of the code (included within the .Rmd file)
- **activity.csv**: The data file used for the analysis, also available [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).
- **activity.zip**: The zipped version of the .csv file




<a name="Code"/>
## Summary of Code and Analysis

#### Loading and preprocessing the data

First, the data is loaded with the read.csv(). The header (first ten lines) are examined to ensure that the data was loaded properly and looks as expected.

```{r,echo = TRUE}
activity=read.csv('activity.csv')
head(activity)
```

We will do any processing of the data in the following steps for specific tasks as necessary.  
<br /><br />


#### What is mean total number of steps taken per day?

Now, we find the mean number of steps taken on each day and examine the distribution by using a histogram.

```{r Plot1 - Histogram,fig.width=5, fig.height=4,echo = TRUE}
datemeans=tapply(activity$steps,activity$date,mean,na.rm=TRUE)
hist(datemeans,col="green",xlab="Mean number of steps per day", ylab="Frequency",main="Distribution of Mean Steps Per Day")
```

We can look at the mean and median of the overall distribution of steps per day:
```{r}
paste ("Mean of total number of steps per day = ",round((mean(datemeans,na.rm=TRUE)),3))
paste ("Median of total number of steps per day = ",round((median(datemeans,na.rm=TRUE)),3))
```
<br /><br />


### What is the average daily activity pattern?

By comparing the number of steps in the same time interval from day to day, we can get an idea of which times of the day, on average, the subject is more or less active.

```{r Plot2 - Mean Steps,fig.width=5, fig.height=4,echo = TRUE}
intervalmeans=tapply(activity$steps,activity$interval,mean,na.rm=TRUE)
intervals=unique(unlist(activity$interval))
plot(intervals,intervalmeans,type='l',col='blue',xlab="Interval",ylab="Mean Steps",main="Time series: Mean number of steps per interval")
```

The peak of this time-series average is at the interval 835, which is the 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps.

```{r}
intervals[intervalmeans==max(intervalmeans,na.rm=TRUE)]
```
<br /><br />


#### Imputing missing values

First, we want to determine how many missing values exist in the dataset:
```{r,fig.width=5, fig.height=4,echo = TRUE}
sum(is.na(activity$steps))
```


Next, we want to fill in these missing values with reasonable values using a logical strategy. Here we employ a technique to fill in the missing data with the mean of that particular 5-minute interval average using the entire dataset. We use this to create a new dataset that is equal to the original dataset, except with the missing data filled in where there were previously "NA"s.

```{r,fig.width=5, fig.height=4,echo = TRUE}
newfile=activity
missing=is.na(newfile$steps)
for(i in 1:length(newfile$steps)){
  if(missing[[i]]==TRUE){
    newfile$steps[[i]]=intervalmeans[as.numeric(names(intervalmeans))==newfile$interval[[i]]]
  }
}
head(newfile)
```
Note that now in the first ten lines we have the average number of steps in the intervals replacing what used to be NA values (see initial header at beginning of file for comparison).

Now we make a histogram of the total number of steps taken each day, similar to the previous histogram, except including the imputed data. 

```{r Plot3 - Histogram 2,fig.width=5, fig.height=4,echo = TRUE}
datemeans_new=tapply(newfile$steps,newfile$date,mean)
hist(datemeans_new,col="green",xlab="Mean number of steps per day", ylab="Frequency",main="Distribution of Mean Steps Per Day")
```

Below we give the mean and median total number of steps taken per day.

```{r,fig.width=5, fig.height=4,echo = TRUE}
paste ("Mean of total number of steps per day = ",round((mean(datemeans_new)),3))
paste ("Median of total number of steps per day = ",round((median(datemeans_new)),3))
```

These values are very similar to those before imputing the data. The impact of the imputing technique used seems to push the mean and median values towards the mean value of all the steps. This does make sense, because we are using the mean at each interval to fill in the missing data.

<br /><br />



#### Are there differences in activity patterns between weekdays and weekends?

After using "strptime" to convert the "date" column into a readable date for R, we determine the day of the week of each date with the weekdays() function. Then we create a new column (newfile$dayofweek) in the new datatable with the imputed values (newfile) that is a factor variable flagging whether it is a "weekday" or "weekend".

```{r,fig.width=5, fig.height=4,echo = TRUE}
times=strptime(newfile$date,format="%Y-%m-%d")
newfile$daynames=weekdays(times)
newfile$dayofweek=ifelse(newfile$daynames %in% c("Saturday","Sunday"),"weekend", "weekday")
```


As before we construct a time-series plot, but now we split based on the factor variable "dayofweek" to compare the weekday time-series number steps to that of the weekend.

```{r Plot4 - Mean Steps: Weekday vs. Weekend,fig.width=8, fig.height=10,echo = TRUE}
weekdaymean=tapply(newfile$steps[newfile$dayofweek=="weekday"],newfile$interval[newfile$dayofweek=="weekday"],mean)
weekendmean=tapply(newfile$steps[newfile$dayofweek=="weekend"],newfile$interval[newfile$dayofweek=="weekend"],mean)

par(mfrow=c(2,1))
plot(intervals,weekdaymean, xlab="Mean number of steps per weekday", ylab="Mean Number of Steps",main="Weekday Activity",type="l",col="red")
plot(intervals,weekendmean, xlab="Mean number of steps per weekend-day", ylab="Mean Number of Steps", main="Weekend Acitivity",type="l",col="magenta")
```

It looks as though, on average, weekend days involve more steps through the middlemost intervals than weekdays do.

<br /><br />



<a name="Conclusions"/>
## Conclusions

To summarize, we examine the data from an activity monitoring device to help understand daily/weekly patterns of its user. Some data is missing and is imputed from the data available. It appears that weekday mornings are quite active for a burst of time. Then, activity is quiet for the majority of the remainder of the day (presumably because the user is at work), though with some additional activity later in the day. Weekend mornings tend to average fewer steps than weekday morning peaks, but the weekends tend to have more activity overall spread throughout the day.




<a name="Other"/>
## Additional Information

Author: RWK

License: None, free to use and edit as people wish.

Contact: May be made through GitHub

<br /><br />



